---
title: "DT 607---Fall 2019---Project 4"
author: "Team ADMJ"
date: "10/27/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r loadLibraries, message=FALSE}
library(tm)
library(SnowballC)
library(data.table)
setDTthreads(0L)
```

# Assignment
It can be useful to be able to classify new "test" documents using already
classified "training" documents.  A common example is using a corpus of labeled
spam and ham (non-spam) e-mails to predict whether or not a new document is spam.  

For this project, you can start with a spam/ham dataset, then predict the class
of new documents (either withheld from the training dataset or from another
source such as your own spam folder). One example corpus:
https://spamassassin.apache.org/old/publiccorpus/

# Solution
## Notes
### Headers
Header should not have blank lines per
[RFC 5322, Section 2.2](https://tools.ietf.org/html/rfc5322). So look for first
blank line.

Remove headers since they will dominate the bag (Received: etc.)

# List files
```{r}
sp_files <- list.files("./Data/spam_2", full.names = TRUE)
h_files <- list.files("./Data/easy_ham", full.names = TRUE)
```

# Build corpus
```{r}
build_Corpus <- function(files, class){
  for (i in seq_along(files)) {
    raw_text <- readLines(files[i])
    em_length <- length(raw_text)
    body_start <- min(grep("^$", raw_text, fixed = FALSE)) + 1L
    em_body <- paste(raw_text[body_start:em_length],
                     sep = "", collapse = " ")
    if (i == 1L) {
      ret_Corpus <- VCorpus(VectorSource(em_body))
    } else {
      tmp_Corpus <- VCorpus(VectorSource(em_body))
      ret_Corpus <- c(ret_Corpus, tmp_Corpus)
    }
  }
  meta(ret_Corpus, tag = "class", type = "indexed") <- class
  return(ret_Corpus)
}
```


# Clean Corpus
```{r}
clean_Corpus <- function(corpus){
  corpus <- tm_map(corpus, removePunctuation,
                   preserve_intra_word_contractions = TRUE,
                   preserve_intra_word_dashes = TRUE)
  corpus <- tm_map(corpus, removeNumbers)
  corpus <- tm_map(corpus, content_transformer(tolower))
  corpus <- tm_map(corpus, removeWords, stopwords("english"))
  corpus <- tm_map(corpus, stemDocument)
  corpus <- tm_map(corpus, stripWhitespace)
  return(corpus)
}
```

